Optimizing Abstract Abstract Machines

### 2013-07-02 paper/paper.tex
Removed the convoluted abstract compilation example and referred the reader back to the state graphs instead. One citation over 12 pages

The possible traces of execution tell us the most about a program, so
we take $\traces(\mexp)$ to be the (regular) set of paths through the
computed graph. We elide the construction of the set of edges in this paper.

To conduct this \naive{} exploration on the \Church{} example would require
considerable time.  Even though the state space is finite, it is exponential in
the size of the program.  Even with $k = 0$, there are exponentially many
stores in the AAM framework.

In the next subsection, we fix this with store widening to reach polynomial
(albeit of high degree) complexity.
%
This widening effectively lifts the store out of individual states to create
a single, global shared store for all.

@@

%%
The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
Now instead of storing the evaluated function in the continuation
frame itself, we indirect it through the store for further control on
complexity and precision:
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr) \\
\text{ where } \maddr &= \alloc(\mstate) \\
               \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}

Associated with this indirection, we now apply all functions stored in
the address. This nondeterminism is necessary in order to continue with
evaluation.

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, pre-allocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics;
 \item lazy non-determinism; and
 \item abstract compilation.
\end{enumerate}

All pragmatic improvements are precision preserving (form complete
abstractions), but the semantic changes are not in some cases, for reasons we
will describe. We did not observe the precision differences in our evaluation.

We apply the frontier-based semantics combined with timestamped stores
as our first step.  The move to the imperative will be made last in
order to show the effectiveness of these techniques in the purely
functional realm.

@@

The semantics given for store widening in section \ref{sec:storewiden},
while simple, is wasteful. It also does not model what typical
implementations do. It causes all states found so far to step each
iteration, even if they are not revisited. This has negative
performance \emph{and} precision consequences (changes to the store
can travel back in time in straight-line code). We instead use a
frontier-based semantics that corresponds to the classic worklist
algorithms for analysis. The difference is that the store is not
modified in-place, but updated after all frontier states have been
processed. This has implications for the analysis' precision and
determinism. Specifically, higher precision, and it is deterministic
even if set iteration is not.

The state space changes from a store and set of contexts to a set
of seen abstract states (context plus store), $\mseen$, a set of contexts to
step (the frontier), $F$, and a store to step those contexts with, $\msto$:
\begin{equation*}
(\mseen, F, \msto) \mathrel{\widehat{\machstep}} (\mseen \cup \mseen', F', \msto')
\end{equation*}

@@

\begin{enumerate}
\item{In our semantics, the store does not change until the entire
    frontier has been explored. This avoids cross-branch pollution
    which would otherwise happen in Shivers' semantics, e.g., when one branch writes to
    address $\maddr$ and another branch reads from address
    $\maddr$.}
\item{The common implementation strategy for timestamps destructively
    updates each state's timestamp. This loses \emph{temporal}
    information about the contexts a state is visited in, and in what
    order. Our semantics has a drop-in replacement of timestamps for
    stores in the seen set ($\mseentime$), so we do not experience precision
    loss.}
\end{enumerate}

@@

The observation Shivers made was that the store is increasing
monotonically, so all stores throughout execution will be totally
ordered (form a chain). This observation allows you to replace stores
with pointers into this chain. We keep the stores around in $\Sigma$
to achieve a complete abstraction. This corresponds to the temporal
information about the execution's effect on the store.

Note also that $F$ is only populated with states that have not been
seen at the resulting store. This is what produces the more precise
abstraction than the baseline widening.

The general fixed-point combinator we showed above can be specialized
to this semantics, as well. In fact, $\widehat{\machstep}^T$ is a functional
relation, so we can get the least fixed-point of it directly.

@@

\begin{theorem}
$\widehat{\machstep}^T$ is a complete abstraction of $\widehat{\machstep}$.
\end{theorem}
The proof follows from the order isomorphism that, in one direction,
sorts all the stores in $\mseen$ to form $\Sigma$, and translates stores in
$\mseen$ to their distance from the end of $\Sigma$ (their timestamp). In the other
direction, timestamps in $\mseentime$ are replaced by the stores they point to in $\Sigma$.

@@

The above technique requires joining entire (large) stores
together. Additionally, there is still a comparison of stores, which
we established is expensive. Not every step will modify all addresses
of the store, so joining entire stores is wasteful in terms of memory
and time. We can instead log store changes and replay the change log
on the full store after all steps have completed, noting when there is
an actual change. This uses far fewer join and comparison operations,
leading to less overhead, and is precision-preserving.

We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times
  \Set(\Storeable))^*$. Each $\msto\sqcup[\maddr \mapsto \mval{s}]$
 becomes a log addition
$\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins
empty ($\mtlst$) for each step. Applying the changes to the full store
is straightforward:
\begin{equation*}
\replay : (\StoreDelta \times \Store) \to (\Store \times \Boolean)
\end{equation*}
\begin{align*}
\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &=
\ttuple{\msto'}
       {\diffp(\mval{s_i}, \msto(\maddr_i)) \vee \ldots} \\
\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots} \\
\diffp(\mval{s}, \mval{s'}) &= \mval{s'} \overset{?}{=} \mval{s} \sqcup \mval{s'}
\end{align*}

We change the semantics slightly to add to the change log rather than
produce an entire modified store.  The transition relation is
identical except for the addition of this change log.  We maintain the
invariant that lookups will never rely on the change log, so we can
use the originally supplied store unmodified.

@@

\spchoice
{
\begin{theorem}[Soundness]
  If $\mstate \machstep \mstate'$ and $\mstate \sqsubseteq
  \mastate$ then there exists a $\mastate'$ such that $\mastate
  \lmachstep \mastate'$ and $\mstate' \sqsubseteq \mastate'$
\end{theorem}
Here $\sqsubseteq$ is straightforward --- the left-hand side store must be
contained in the right-hand-side store, and if values occur in the states,
the left-hand-side value must be in the forced corresponding right-hand-side value.
The proof is by cases on $\mstate \machstep \mstate'$.}
{
\begin{theorem}[Completeness]For all $\mexp$,
 $\traces_{\text{Lazy}}(\mexp)$ is a complete abstraction of $\traces_{\text{ISWIM}}(\mexp)$.
\end{theorem}
We have a statement about traces because we need induction to show no
cruft values are in superposition.  The induction hypothesis tells us
that there are non-lazy traces that lead to all the values in
superposition, so when we take a lazy step, we are taking several
non-lazy steps, and we stay in sync. The other direction we just
collapse the superposition in each possibility to construct the
non-lazy traces.}

@@

The prior optimization saved time by doing the same amount of
reasoning as before but in fewer transitions. We can exploit the same
idea---same reasoning, fewer transitions---with abstract
compilation. Abstract compilation transforms complex expressions whose
\emph{abstract} evaluation is deterministic into ``abstract
bytecodes.''  The abstract interpreter then does in one transition
what previously took many. In short, abstract compilation eliminates
unnecessary allocation, deallocation and branching. The technique is
precision preserving without store widening. We discuss the precision
differences with store widening at the end of the section.

The following example illustrates
the essence of abstract compilation's effect:
\[
\mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
\]
makes the following transitions:
\begin{align}
& \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
\machstep\; &
\ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_1}}
\\
\machstep\; &
\ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_2}}
\\
\machstep\; &
\ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_3}} % {\mexp_2}
\\
\machstep\; &
\spchoice
{\co{\kar{\mexp_1,\menv,\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
{\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
\end{align}
where \(
\msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
\maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
\maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).

Whereas abstract compilation gives us in one step:
\begin{align*}
\compile{\mexp}^\mcntr(\msto_0)(\menv,\epsilon,\mkont) =
\spchoice
{\co{\kar{\compile{\mexp_1},\menv,\maddr_3},\saddr{\menv(\mvar)}},\msdiff_4}
{\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
\end{align*}
where
\begin{equation*}
\msdiff_4 = [\ttuple{\maddr_1}{\setof{\mkont}},
             \ttuple{\maddr_2}{\setof{\kar{\compile{\mexp_3},\menv,\maddr_1}}},
             \ttuple{\maddr_3}{\setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}}]\text.
\end{equation*}

The compilation step converts expressions into functions that expect
the other components of the {\tt ev} state. Its definition in figure
\ref{fig:compile} shows close similarity to the rules for interpreting
{\tt ev} states. The next step is to change reduction rules that
create {\tt ev} states to instead call these functions. Figure
\ref{fig:caam} shows the modified reduction relation. The only change
from the previous semantics is that $\evalone$ state construction is
replaced by calling the compiled expression. For notational coherence,
we write $\lambda^\mcntr(\mathit{args} \ldots)$ for
$\lambda(\mathit{args} \ldots, \mcntr)$ and
$\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
\mcntr)$.

>>>>

The possible traces of execution tell us the most about a program, so we take $\traces(\mexp)$ to be the (regular) set of paths through the computed graph.
%
 We elide the construction of the set of edges in this paper.
%%

%%
To conduct this \naive{} exploration on the \Church{} example would require considerable time.
%
  Even though the state space is finite, it is exponential in the size of the program.
%
  Even with $k = 0$, there are exponentially many stores in the AAM framework.
%%

%%
In the next subsection, we fix this with store widening to reach polynomial (albeit of high degree) complexity.
%
This widening effectively lifts the store out of individual states to create a single, global shared store for all.

@@

%%
The new evaluation rules follow, where $\mcntr' = \tick(\mstate)$:
% In theory, this aggressive distinction between continuations might buy
% additional precision, but in practice, it does not.
\begin{align*}
\co[^\mcntr]{\kar{\mexp,\menv,\mkaddr},\mval,\msto} & \machstep
\ev[^{\mcntr'}]{\mexp,\menv,\msto',\kfn{\maddr,\mkaddr}} \\
\text{ where }
  \maddr &= \alloc(\mstate) \\
  \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
%%
Now instead of storing the evaluated function in the continuation frame itself, we indirect it through the store for further control on complexity and precision:
%%
\begin{align*}
\co[^\mcntr]{\kfn{\maddr,\mkaddr},\mval,\msto} & \machstep
\ap[^{\mcntr'}_\mlab]{\mvalx{u},\maddr,\mkont,\msto'}
\\
\text{ if } \mkont &\in \msto(\mkaddr), \mvalx{u} \in \msto(\maddr) \\
\text{ where } \maddr &= \alloc(\mstate) \\
               \msto' &= \ext{\msto}{\maddr}{\{\mval\}}
\end{align*}
%%

%%
Associated with this indirection, we now apply all functions stored in the address.
%
This nondeterminism is necessary in order to continue with evaluation.

@@

In this section, we discuss the optimizations for abstract interpreters that
yield our ultimate performance gains.
%
We have two broad categories of these optimizations: (1) pragmatic
improvement, (2) transition elimination.
%
The pragmatic improvements reduce overhead and trade space for time
by utilizing:
\begin{enumerate}
 \item timestamped stores;
 \item store deltas; and
 \item imperative, pre-allocated data structures.
\end{enumerate}
The transition-elimination optimizations reduce the overall number of transitions
made by the analyzer by performing:
\begin{enumerate}
  \setcounter{enumi}{3}
 \item frontier-based semantics;
 \item lazy non-determinism; and
 \item abstract compilation.
\end{enumerate}
%%

%%
All pragmatic improvements are precision preserving (form complete abstractions), but the semantic changes are not in some cases, for reasons we will describe.
%
 We did not observe the precision differences in our evaluation.
%%

%%
We apply the frontier-based semantics combined with timestamped stores as our first step.
%
  The move to the imperative will be made last in order to show the effectiveness of these techniques in the purely functional realm.

@@

The semantics given for store widening in section \ref{sec:storewiden}, while simple, is wasteful.
%
 It also does not model what typical implementations do.
%
 It causes all states found so far to step each iteration, even if they are not revisited.
%
 This has negative performance \emph{and} precision consequences (changes to the store can travel back in time in straight-line code).
%
 We instead use a frontier-based semantics that corresponds to the classic worklist algorithms for analysis.
%
 The difference is that the store is not modified in-place, but updated after all frontier states have been processed.
%
 This has implications for the analysis' precision and determinism.
%
 Specifically, higher precision, and it is deterministic even if set iteration is not.
%%

%%
The state space changes from a store and set of contexts to a set of seen abstract states (context plus store), $\mseen$, a set of contexts to step (the frontier), $F$, and a store to step those contexts with, $\msto$:
\begin{equation*}
(\mseen, F, \msto) \mathrel{\widehat{\machstep}} (\mseen \cup \mseen', F', \msto')
\end{equation*}

@@

\begin{enumerate}
\item{In our semantics, the store does not change until the entire frontier has been explored.
%
 This avoids cross-branch pollution which would otherwise happen in Shivers' semantics, e.g., when one branch writes to address $\maddr$ and another branch reads from address $\maddr$.}
\item{The common implementation strategy for timestamps destructively updates each state's timestamp.
%
 This loses \emph{temporal} information about the contexts a state is visited in, and in what order.
%
 Our semantics has a drop-in replacement of timestamps for stores in the seen set ($\mseentime$), so we do not experience precision loss.}
\end{enumerate}

@@

The observation Shivers made was that the store is increasing monotonically, so all stores throughout execution will be totally ordered (form a chain).
%
 This observation allows you to replace stores with pointers into this chain.
%
 We keep the stores around in $\Sigma$ to achieve a complete abstraction.
%
 This corresponds to the temporal information about the execution's effect on the store.

Note also that $F$ is only populated with states that have not been seen at the resulting store.
%
 This is what produces the more precise abstraction than the baseline widening.

The general fixed-point combinator we showed above can be specialized to this semantics, as well.
%
 In fact, $\widehat{\machstep}^T$ is a functional relation, so we can get the least fixed-point of it directly.

@@

\begin{theorem}
$\widehat{\machstep}^T$ is a complete abstraction of $\widehat{\machstep}$.
\end{theorem}
The proof follows from the order isomorphism that, in one direction, sorts all the stores in $\mseen$ to form $\Sigma$, and translates stores in $\mseen$ to their distance from the end of $\Sigma$ (their timestamp).
%
 In the other direction, timestamps in $\mseentime$ are replaced by the stores they point to in $\Sigma$.

@@

The above technique requires joining entire (large) stores together.
%
 Additionally, there is still a comparison of stores, which we established is expensive.
%
 Not every step will modify all addresses of the store, so joining entire stores is wasteful in terms of memory and time.
%
 We can instead log store changes and replay the change log on the full store after all steps have completed, noting when there is an actual change.
%
 This uses far fewer join and comparison operations, leading to less overhead, and is precision-preserving.

We represent change logs as $\msdiff \in \StoreDelta = (\Addr \times \Set(\Storeable))^*$.
%
 Each $\msto\sqcup[\maddr \mapsto \mval{s}]$ becomes a log addition $\cons{\ttuple{\maddr}{\mval{s}}}{\msdiff}$, where $\msdiff$ begins empty ($\mtlst$) for each step.
%
 Applying the changes to the full store is straightforward:
\begin{equation*}
\replay : (\StoreDelta \times \Store) \to (\Store \times \Boolean)
\end{equation*}
\begin{align*}
\replay(\left[ \ttuple{\maddr_i}{\mval{s_i}}, \ldots\right], \msto) &=
\ttuple{\msto'}
       {\diffp(\mval{s_i}, \msto(\maddr_i)) \vee \ldots} \\
\text{ where } \msto' &= {\msto \sqcup [\maddr_i \mapsto \mval{s_i}] \sqcup \ldots} \\
\diffp(\mval{s}, \mval{s'}) &= \mval{s'} \overset{?}{=} \mval{s} \sqcup \mval{s'}
\end{align*}

We change the semantics slightly to add to the change log rather than produce an entire modified store.
%
The transition relation is identical except for the addition of this change log.
%
  We maintain the invariant that lookups will never rely on the change log, so we can use the originally supplied store unmodified.

@@

\spchoice
{
\begin{theorem}[Soundness]
  If $\mstate \machstep \mstate'$ and $\mstate \sqsubseteq \mastate$ then there exists a $\mastate'$ such that $\mastate \lmachstep \mastate'$ and $\mstate' \sqsubseteq \mastate'$
\end{theorem}
Here $\sqsubseteq$ is straightforward --- the left-hand side store must be contained in the right-hand-side store, and if values occur in the states, the left-hand-side value must be in the forced corresponding right-hand-side value.
%
The proof is by cases on $\mstate \machstep \mstate'$.}
{
\begin{theorem}[Completeness]For all $\mexp$,
 $\traces_{\text{Lazy}}(\mexp)$ is a complete abstraction of $\traces_{\text{ISWIM}}(\mexp)$.
\end{theorem}
We have a statement about traces because we need induction to show no cruft values are in superposition.
%
The induction hypothesis tells us that there are non-lazy traces that lead to all the values in superposition, so when we take a lazy step, we are taking several non-lazy steps, and we stay in sync.
%
 The other direction we just collapse the superposition in each possibility to construct the non-lazy traces.}

@@

The prior optimization saved time by doing the same amount of reasoning as before but in fewer transitions.
%
 We can exploit the same idea---same reasoning, fewer transitions---with abstract compilation.
%
 Abstract compilation transforms complex expressions whose \emph{abstract} evaluation is deterministic into ``abstract bytecodes.''
%
  The abstract interpreter then does in one transition what previously took many.
%
  Refer back to figure \ref{fig:state-graphs} to see the effect of abstract compilation.
%
 In short, abstract compilation eliminates unnecessary allocation, deallocation and branching.
%
 The technique is precision preserving without store widening.
%
 We discuss the precision differences with store widening at the end of the section.
%%

% %5
% The following example illustrates
% the essence of abstract compilation's effect:
% \[
% \mexp := \sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3}
% \]
% makes the following transitions:
% \begin{align}
% & \ev[^\mcntr]{\sapp{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2}}{\mexp_3},\menv,\msto_0,\mkont}\\
% \machstep\; &
% \ev[^{\mcntr'}]{\sapp{\sapp\mvar{\mexp_1}}{\mexp_2},\menv,\msto_1,\kar{\mexp_3,\menv,\maddr_1}}
% \\
% \machstep\; &
% \ev[^{\mcntr''}]{\sapp\mvar{\mexp_1},\menv,\msto_2,\kar{\mexp_2,\menv,\maddr_2}}
% \\
% \machstep\; &
% \ev[^{\mcntr'''}]{\mvar, \menv,\msto_3,\kar{\mexp_1,\menv,\maddr_3}} % {\mexp_2}
% \\
% \machstep\; &
% \spchoice
% {\co{\kar{\mexp_1,\menv,\maddr_3},\saddr{\menv(\mvar)},\msto_4}}
% {\co{\kar{\mexp_1,\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto_4(\menv(\mvar))},\msto_4}}
% \end{align}
% where \(
% \msto_4 = \msto_0 \sqcup [\maddr_1 \mapsto \setof{\mkont},
% \maddr_2 \mapsto \setof{\kar{\compile{\mexp_3},\menv,\maddr_1}},
% \maddr_3 \mapsto \setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}]\).

% Whereas abstract compilation gives us in one step:
% \begin{align*}
% \compile{\mexp}^\mcntr(\msto_0)(\menv,\epsilon,\mkont) =
% \spchoice
% {\co{\kar{\compile{\mexp_1},\menv,\maddr_3},\saddr{\menv(\mvar)}},\msdiff_4}
% {\co{\kar{\compile{\mexp_1},\menv,\maddr_{f3},\maddr_{a3},\maddr_3},\superposition{\msto'_4(\menv(\mvar))},\msto'_4}}
% \end{align*}
% where
% \begin{equation*}
% \msdiff_4 = [\ttuple{\maddr_1}{\setof{\mkont}},
%              \ttuple{\maddr_2}{\setof{\kar{\compile{\mexp_3},\menv,\maddr_1}}},
%              \ttuple{\maddr_3}{\setof{\kar{\compile{\mexp_2},\menv,\maddr_2}}}]\text.
% \end{equation*}

The compilation step converts expressions into functions that expect the other components of the {\tt ev} state.
%
Its definition in figure \ref{fig:compile} shows close similarity to the rules for interpreting {\tt ev} states.
%
 The next step is to change reduction rules that create {\tt ev} states to instead call these functions.
%
 Figure \ref{fig:caam} shows the modified reduction relation.
%
 The only change from the previous semantics is that $\evalone$ state construction is replaced by calling the compiled expression.
%
 For notational coherence, we write $\lambda^\mcntr(\mathit{args} \ldots)$ for $\lambda(\mathit{args} \ldots, \mcntr)$ and $\mcomp^\mcntr(\mathit{args}\ldots)$ for $\mcomp(\mathit{args}\ldots,
\mcntr)$.
